---
title: "M4_W2_L2_ggplot2"
author: "Alexander Cormack"
date: "2022-11-19"
output: html_document
---

# ggplot2

## What is ggplot2?

An implementation of the *Grammar of Graphics* by Leland Wilkinson

Written by Hadley Wickham (while he was a graduate student at Iowa State)

A "third" graphics system for R (along with **base** and *lattice**)

Available from CRAN via install.packages()

Web site: http:/ggplot2.org (better documentation)

Grammar of graphics represents an abstraction of graphics ideas/objects

Think "verb", "noun", "adjective" for graphics

Allows for a "theory" of graphics on which to build new graphics and graphics objects

"Shorten the distance from mind to page"

*In brief, the grammar tells us that a statistical graphic is a* **mapping** *from data to* **aesthetic** *attributes (colour, shape, size) of* **geometric** *objects (points, lines, bars). The plot may also contain statistical transformations of the data and is drawn on a specific coordinate system*
(from *ggplot2* book)


## The basics: qplot()

Works much like the plot function in base graphics system

Looks for data in a data frame, similar to lattice, or in the parent environment

Plots are made up of *aesthetics* (size, shape, colour) and *geoms* (points, lines)

Factors are important for indicating subsets of the data (if they have different properties); they should be **labeled**

The qplot() hides what goes on underneath, which is okay for most operations

ggplot() is the core function and very flexible for doing things qplot() cannot do


```{r}
library(ggplot2)
str(mpg)
```
```{r}
qplot(displ, hwy, data = mpg)
```

### Modifying aesthetics

```{r}
qplot(displ, hwy, data = mpg, color = drv)
```

We can add a statistic showing the general trend - the smoother is the blue line and the 95% confidence intervals are represented by the gray zone

```{r}
qplot(displ, hwy, data = mpg, geom = c("point", "smooth"))
```

We can also create a histogram by specifying a single variable

```{r}
qplot(hwy, data = mpg, fill = drv)
```

### Facets

Facets are like panels in lattice. If you have a lot of data points that you want to separate out, using the colour scheme above may not be the best solution. It may be easier to separate them with facets.

The facets function takes two arguments separated by a tilde. The argument on the left represents the rows, the argument on the right the columns.

In the examples below the . indicates there is no other argument so there is only one row in the first plot and one column in the second plot.

```{r}
qplot(displ, hwy, data = mpg, facets = .~ drv)
```

```{r}
qplot(hwy, data = mpg, facets = drv ~., binwidth = 2)
```


## Summary of qplot()

The qplot() function is the analog to plot() but with many built-in features

Syntax is somewhere between base/lattuce

Produces nice graphics, essentially publication ready (if you like the design)

Difficult to go against the grain/customise (don't bother - use ggplot2 power in that case)












