---
title: "M4_W3_L3_Working_With_Colour_In R_Plots"
author: "Alexander Cormack"
date: "2022-12-04"
output: html_document
---

# Working With Colours In R Plots


## Plotting and colour in R 

- The default colour schemes for most plots in R are horrendous

- Recently [sic] there have been developments to improve the handling / specification of colours in plots / graphs / etc.

- There are functions in R and in external packages that are very handy


## Colour utilities in R

- The **grDevices** package has two functions
    - colorRamp
    - colorRampPalette

- These functions take palettes of colours and help to interpolate between the colours

- The function colors() lists the names of colours you can use in any plotting function


## Colour palette utilities in R

- colorRamp: take a palette of colours and return a function that takes values between 0 and 1, indicating the extremes of the colour palette (e.g., see the 'gray' function)

- colorRampPalette: take a palette of colours and return a function that takes integer arguments and returns a vector of colours interpolating the palette (like heat.colors or topo.colors)


### colorRamp

```{r}
pal <- colorRamp(c("red", "blue"))
pal(0)
```
```{r}
pal(1)
```
```{r}
pal(0.5)
```
```{r}
pal(seq(0, 1, len = 10))
```

### colorRampPalette

```{r}
pal <- colorRampPalette((c("red", "yellow")))
pal(2)
```
```{r}
pal(10)
```


## RColorBrewer package

- One package on CRAN that contains interesting/useful colour palettes

- There are three types of palettes:
    - sequential
    - diverging
    - qualitative


- Palette information can be used in conjunction with the colorRamp() and colorRampPalette() functions
  
  
### RColorBrewer and colorRampPalette

```{r}
library(RColorBrewer)
cols <- brewer.pal(3, "BuGn")
cols
```
```{r}
pal <- colorRampPalette(cols)
image(volcano, col = pal(20))
```

### The smoothScatter function

```{r}
x <- rnorm(10000)
y <- rnorm(10000)
smoothScatter(x, y)
```


## Some other plotting notes

- The rgb function can be used to produce any colour via proportions of red, green and blue

- Colour transparency can be added via the alpha parameter to rgb

- The **colorspace** package can be used for a different control over colours


### Scatterplot with no transparency

```{r}
x <- rnorm(2000)
y <- rnorm(2000)
plot(x, y, pch = 19)
```

### Scatterplot with tranparency

```{r}
plot(x, y, col = rgb(0, 0, 0, 0.2), pch = 19)
```


## Summary

- Careful use of colours in plots/maps/etc. can make it easier for the reader to get what you're trying to say (why make it harder?)

- The **RColorBrewer** package is an R package that provides colour palettes for sequential, categorical and diverging data

- The colorRamp and colorRampPalette functions can be used in conjunction with colour palettes to connect data to colours

- Transparency can sometimes be used to clarify plots with many points











