---
title: "M4_W1_L2_Plotting"
author: "Alexander Cormack"
date: "7 November 2022"
output: html_document
---


# Plotting Systems

## The base plotting system

The conceptual model used by the base plotting system is the "Artist's Palette" model, i.e. you start with a blank canvas and then you build from there. Every little piece of the plot is added to the plot one by one.

This is fine if you are keeping track of the code for the plot so you can re-plot it again later.

Generally you start with a plot function or similar and then you use annotation functions to add or modify elements of the plot (e.g. text, lines, points, axes)

This approach is quite convenient, as it mirrors how we think of building plots and analysing data.

One of the drawbacks though is that you can't go back once the plot has started (e.g. to adjust the margins); you need to plan in advance.

It is difficult to "translate" to others once a new plot has been created as there is no graphical "language".

Lastly, the plot is just a series of R commands


Here we have a very simple base plot using the cars dataset - it shows the distance it takes a car to stop given a certain speed.

```{r}
library(datasets)
data(cars)
with(cars, plot(speed, dist))
```

From here you could do a lot of things, e.g. title, labels, different colored points, shapes, etc.


## The lattice system

- Plots are created with a single function call (xyplot, bwplot, etc.)
- Most useful for conditioning types of plots: looking at how y changes with x across levels of z
- Things like margins/spacing set automatically bacause entire plot is specified at once
- Good for putting many many plots on a screen

Some downsides

- Sometimes awkward to specify an entire plot in a single function call
- Annotation in plot is not especially intuitive
- Use of panel functions and subscripts difficult to wield and requires intense preparation
- Cannot "add" to the plot once it is created

```{r}
library(lattice)
state <- data.frame(state.x77, region = state.region)
xyplot(Life.Exp ~ Income | region, data = state, layout = c(4, 1))
```


## The ggplot2 system

- Splits the difference between base and lattice in a number of ways
- Automatically deals with spacings, text, titles but also allows you to annotate by "adding" to a plot
- Superficial similarity to lattice but generally easier/more intuitive to use
- Default mode makes many choices for you (but you can still customise to your heart's desire)


Here is a simple ggplot2 graph using the miles per gallon dataset included in the library. It shows the displacement or size of an engine on the x axis and the mileage on the highway on the y axis. You can see that some of the default features are different from the other two systems.

```{r}
library(ggplot2)
data(mpg)
qplot(displ, hwy, data = mpg)
```


## Summary

There are three basic plotting systems in R:

- Base: "artist's palette" model
- Lattice: entire plot specified by one function; conditioning
- ggplot2: mixes elements of Base and Lattice

















