---
title: "M4_W1_L2_Plotting"
author: "Alexander Cormack"
date: "7 November 2022"
output: html_document
---


# Plotting Systems

## The base plotting system

The conceptual model used by the base plotting system is the "Artist's Palette" model, i.e. you start with a blank canvas and then you build from there. Every little piece of the plot is added to the plot one by one.

This is fine if you are keeping track of the code for the plot so you can re-plot it again later.

Generally you start with a plot function or similar and then you use annotation functions to add or modify elements of the plot (e.g. text, lines, points, axes)

This approach is quite convenient, as it mirrors how we think of building plots and analysing data.

One of the drawbacks though is that you can't go back once the plot has started (e.g. to adjust the margins); you need to plan in advance.

It is difficult to "translate" to others once a new plot has been created as there is no graphical "language".

Lastly, the plot is just a series of R commands


Here we have a very simple base plot using the cars dataset - it shows the distance it takes a car to stop given a certain speed.

```{r}
library(datasets)
data(cars)
with(cars, plot(speed, dist))
```

From here you could do a lot of things, e.g. title, labels, different colored points, shapes, etc.


## The lattice system

- Plots are created with a single function call (xyplot, bwplot, etc.)
- Most useful for conditioning types of plots: looking at how y changes with x across levels of z
- Things like margins/spacing set automatically bacause entire plot is specified at once
- Good for putting many many plots on a screen

Some downsides

- Sometimes awkward to specify an entire plot in a single function call
- Annotation in plot is not especially intuitive
- Use of panel functions and subscripts difficult to wield and requires intense preparation
- Cannot "add" to the plot once it is created

```{r}
library(lattice)
state <- data.frame(state.x77, region = state.region)
xyplot(Life.Exp ~ Income | region, data = state, layout = c(4, 1))
```


## The ggplot2 system

- Splits the difference between base and lattice in a number of ways
- Automatically deals with spacings, text, titles but also allows you to annotate by "adding" to a plot
- Superficial similarity to lattice but generally easier/more intuitive to use
- Default mode makes many choices for you (but you can still customise to your heart's desire)


Here is a simple ggplot2 graph using the miles per gallon dataset included in the library. It shows the displacement or size of an engine on the x axis and the mileage on the highway on the y axis. You can see that some of the default features are different from the other two systems.

```{r}
library(ggplot2)
data(mpg)
qplot(displ, hwy, data = mpg)
```


## Summary

There are three basic plotting systems in R:

- Base: "artist's palette" model
- Lattice: entire plot specified by one function; conditioning
- ggplot2: mixes elements of Base and Lattice


# Base Plotting System (Part 1)

The core plotting and graphics engine in R is encapsulated in the following packages:

- *graphics*: contains plotting functions for the "base" graphing systems, including **plot, hist, boxplot** and many others
- *grDevices*: contains all the code implmenting the various graphics devices, including X11, PDF, PostScript, PNG, etc.

The lattice plotting system is implemented using the following packages:

- *lattice*: contains code for producing Trellis graphics, which are independent of the "base" graphics system; includes functions like **xyplot, bwplot, levelplot**
- *grid*: implements a different graphing system independent of the "base" system; the *lattice* package builds on top of *grid*; we seldom call functions from the *grid* package directly


## The process of making a plot

When making a plot one must first make a few considerations (not necessarily in this order):

- Where will the plot be made? On the screen? In a file?

- How will the plot be used?
    - Is the plot for viewing temporarily on the screen?
    - Will it be presented in a web browser?
    - Will it eventually end up in a paper that might be printed?
    - Are you using it in a presentation?


- Is there a large amount of data going into the plot? Or is it just a few points?

- Do you need to be able to dynamically resize the graphic?

- What graphics system will you use: base, lattice or ggplot2? These generally cannot be mixed.

- Base graphics are usually constructed piecemeal, with each aspect of the plot handled separately through a series of function calls; this is often conceptually simpler and allows plotting to miror the thought process

- Lattice graphics are usually created in a single function call, so all of the graphics parameters have to be specified at once; specifying everything at once allows R to automatically calculate the necessary spacings and font sizes.

- ggplot2 combines concepts from both base and lattice graphics but uses an independent implementation

We focus on using the **base plotting system** to create graphics on the **screen device**.


## Base graphics

Base graphics are used most commonly and are a very powerful system for creating 2-D graphics.

There are two phases to creating a base plot:

- Initialising a new plot
- Annotating (adding to) an existing plot

Calling **plot(x, y)** or **hist(x)** will launch a graphics device (if one is not already open) and draw a new plot on the device.

If the arguments to **plot** are not of some special class, then the *default* method for **plot** is called; this function has *many* arguments, letting you set the title, x axis label, y axis label, etc.

The base graphics system has *many* parameters that can be set and tweaked; these parameters are documented in **?par**; it wouldn't hurt to try to memorise this help page!


## Simple base graphics

### Histogram

```{r}
library(datasets)
hist(airquality$Ozone, col = "steelblue") ## Draw a new plot
```

### Scatterplot

```{r}
library(datasets)
with(airquality, plot(Wind, Ozone, pch = 2))
```

### Boxplot

```{r}
library(datasets)
airquality <- transform(airquality, Month = factor(Month))
boxplot(Ozone ~ Month, airquality, xlab = "Month", ylab = "Ozone (ppb)")
```

## Some important base graphics parameters

Many base plotting functions share a set of parameters. Here are a few key ones:

- **pch**: the plotting symbol (default is open circle)

- **lty**: the line type (default is solid line), can be dashed, dotted, etc.

- **lwd**: the line width, specified as an integer multiple

- **col**: the plotting colour, specified as a number, string, or hex code; the **colors()** function gives you a vector of colors by name

- **xlab**: character string for the x-axis label

- **ylab**: character string for the y-axis label


The par() function is used to specify *global* graphics parameters that affect all plots in an R session. These parameters can be overridden when specified as arguments to specific plotting functions.

- **las**: the orientation of the axis labels on the plot

- **bg**: the background colour

- **mar**: the margin size

- **oma**: the outer margin size (default is 0 for all sides)

- **mfrow**: the number of plots per row, column (plots are filled row-wise)

- **mfcol**: number of plots per row, column (plots are filled column-wise)


Default values for global graphics paramaters

```{r}
par("lty")
```

```{r}
par("col")
```

```{r}
par("pch")
```

```{r}
par("bg")
```

```{r}
par("mar")
```

```{r}
par("mfrow")
```


# Base Plotting System (Part 2)

## Base plotting functions

- **plot**: make a scatterplot, or other type of plot depending on the class of the object being plotted

- **lines**: add lines to a plot, given a vector of x values and a corresponding vector of y values (or a 2-column matrix); this function just connects the dots

- **points**: add points to a plot

- **text**: add text labels to a plot using specified x, y coordinates

- **title**: add annotations to x, y axis labels, title, subtitle, outer margin

- **mtext**: add arbitrary text to the margins (inner or outer) of the plot

- **axis**: adding axis ticks/labels


## Base plot with annotation

```{r}
library(datasets)
with(airquality, plot(Wind, Ozone))
title(main = "Ozone and Wind in New York City") ## Add a title
```

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City"))
with(subset(airquality, Month == 5), points(Wind, Ozone, col = "blue"))
```

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City", type = "n")) ## type = "n" sets up the plot without any data
with(subset(airquality, Month == 5), points(Wind, Ozone, col = "blue"))
with(subset(airquality, Month != 5), points(Wind, Ozone, col = "red"))
legend("topright", pch = 1, col = c("blue", "red"), legend = c("May", "Other Months"))
```


## Base plot with regression line

```{r}
with(airquality, plot(Wind, Ozone, main = "Ozone and Wind in New York City", pch = 20))
model <- lm(Ozone ~ Wind, airquality)
abline(model, lwd = 2)
```


## Multiple base plots

```{r}
par(mfrow = c(1, 2))
with(airquality, {
        plot(Wind, Ozone, main = "Ozone and Wind")
        plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
})
```

```{r}
par(mfrow = c(1, 3), mar = c(4, 4, 2, 3), oma = c(0, 0, 2, 0))
with(airquality, {
        plot(Wind, Ozone, main = "Ozone and Wind")
        plot(Solar.R, Ozone, main = "Ozone and Solar Radiation")
        plot(Temp, Ozone, main = "Ozone and Temperature")
        mtext("Ozone and Weather in New York City", outer = TRUE)
})
```

## Summary

Plots in the base plotting system are created by calling successive R functions to "build up" a plot

Plotting occurs in two stages:

- creation of a plot
- annotation of a plot

The base plotting system is very flexible and offers a high degree of control over plottiing.


# Base Plotting Demonstration

We are going to play with some synthetic data to try out the base plotting functionalities.

```{r}
x <- rnorm(100)
```
```{r}
hist(x)
```

We can see a number of defualt elements in the histogram ... axis labels and a title.
```{r}
y <- rnorm(100)
```
```{r}
plot(x, y)
```

In this example we can see the default axes labels and the default plotting symbot (the open circle).

If the above plot appeared in the plotting window we would also notice the default margins, i.e. 5 lines of text for the bottom margin, 4 lines of text for the left and top margin, and two lines of text for the right margin.

The margins can be adjusted using the mar function.
```{r}
par(mar = c(2, 2, 2, 2))
plot(x, y)
```

This has made the margins around the plot even, but it has also meant that there is no space available for the axes labels.

Here we change the plotting symbol.
```{r}
plot(x, y, pch = 20)
```

There is a demo file that shows all of the points options

```{r}
example(points)
```

We can add a title, text, a legend or a line to the plot.
```{r}
x <- rnorm(100)
y <- rnorm(100)
plot(x, y)
title("Scatterplot")
text(-2, -2, "Label")
legend("topleft", legend = "Data", pch = 20)
fit <- lm(y ~ x)
abline(fit, lwd = 2, col = "blue")
```

## Multiple scatterplots

```{r}
z <- rpois(100, 2)
par(mfrow = c(1, 2))
plot(x, y, pch = 20)
plot(x, z, pch = 2)
```

The mfrow parameter adds the plots row-wise where as the mfcol parameter adds them column-wise.

```{r}
par(mfrow = c(2, 2))
plot(x, y)
plot(x, z)
plot(z, x)
plot(y, x)
```
```{r}
par(mfcol = c(2, 2))
plot(x, y)
plot(x, z)
plot(z, x)
plot(y, x)
```

Lastly let's look at how you can annotate data points.
```{r}
par(mfrow = c(1, 1))
x <- rnorm(100) # some synthetic data
y <- x + rnorm(100) #some more synthetic data
g <- gl(2, 50, labels = c("Male", "Female")) # a factor variable with two levels
plot(x, y)
```

As it stands we can't really tell which data points are the males and which data points are the females.

So we are going to build an empty plot and add the various elements one at a time.
```{r}
plot(x, y, type = "n")
```
```{r}
plot(x, y, type = "n")
points(x[g == "Male"], y[g == "Male"], col = "green", pch = 19)
```
```{r}
plot(x, y, type = "n")
points(x[g == "Male"], y[g == "Male"], col = "green", pch = 19)
points(x[g == "Female"], y[g == "Female"], col = "blue")
```













