---
title: "M4_W1_L3_Graphics_Devices_in_R"
author: "Alexander Cormack"
date: "2022-11-12"
output: html_document
---

# Graphics Devices In R

## What is a graphics device?

A graphics device is somethinig where you can make a plot appear

- a window on your computer (screen device)
- a PDF file (file device)
- a PNG or JPEG file (file device)
- a scalable vector graphics (SVG) file (file device)

When you make a plot in R it has to be "sent" to a specific graphics device

The most common place for a plot to be "sent" is the *screen device*

- on a MAC the screen device is launched with the **quartz()**
- on Windows the screen device is launched with **windows()**
- on Unix/Linux the screeen device is launched with **x11()**

When making a plot you need to consider how the plot will be used to determine what device the plot should be sent to.

- the list of devices is found in **?Devices**; there are also devices created by users on CRAN

For quick visualisations and exploratory analysis, usually you want to use the screen device.

- functions like **plot** in base, **xyplot** in lattice, or **qplot** in ggplot2 will default to sending a plot to the screen device
- on a given platform (MAC, Windows, Unix/Linux) there is only one screen device

For plots that may need to be printed out or be incorporated into a dcoument (e.g. papers/reports, slide presentations), usually a *file device* is more appropriate.

- there are many different file devices to choose from

NOTE: Not all graphics devices are available on all platforms (i.e. you cannot launch the **windows()** on a MAC)


## How does a plot get created?

There are two basic approaches to plotting. The first is the most common:

1. Call a plotting function like **plot, xyplot** or **qplot*
2. The plot appears on the screen device
3. Annotate the plot if necessary
4. Enjoy
```{r}
library(datasets)
with(faithful, plot(eruptions, waiting)) ## Make plot appear on screen device
title(main = "Old Faithful Geyser data") ## Annotate with a title
```

The second approach to plotting is most commonly used for file devices:

1. Explicitly launch a graphics device
2. Call a plotting function to make a plot (Note: if you are using a file device, no plot will appear on the screen)
3. Annotate the plot if necessary
4. Explicitly close graphics device with dev.off() (this is very important)
```{r}
pdf(file = "./support/myplot.pdf") ## Open PDF device; create "myplot.pdf" in the "support" directory
## Create plot and send to a file (no plot appears on the screen)
with(faithful, plot(eruptions, waiting))
title(main = "Old Faithful Geyser data") ##Annotate plot; still nothing on screen
dev.off() ## Close the PDF file device
## Now you can view the file "myplot.pdf" on your computer
```

## Graphics file devices

There are two basic types of file devices: *vector* and *bitmap* devices

### Vector formats

- **pdf**: useful for line-type graphics, resizes well, usually portable, not efficient if a plot has many objects/points
- **svg**: XML-based scalable vector graphics; supports animation and interactivity, potentially useful for web-based plots
- **win.metafile**: Windows metafile format (only on Windows)
- **postscript**: older format, also resizes well, usually portable, can be used to create encapsulated postscript files; Windows systems often don't have a postscript viewer


### Bitmap formats

- **png**: bitmapped format, good for line drawings or images with solid colours, uses lossless compression (like the GIF format), most web browsers can read this format natively, good for plotting many many many points, does not resize well
- **jpeg**: good for photographs or natural scenes, uses lossy compression, good for plotting many many many points, does not resize well, can be read by almost any computer and any web browser, not great for line drawings
- **tiff**: creates bitmap files in the TIFF fromat; supports lossless compression
- **bmp**: a native Windows bitmapped format


## Multiple open graphics devices

It is possible to open multiple graphics devices (screen, file or both), for example when viewing multiple plots at once

Plotting can only occur on one graphics device at a time

The **currently active** graphics device can be found by calling **dev.cur()**

Every open graphics device is assigned an integer >= 2

You can change the active graphics device with the **dev.set(integer)** where **integer** is the number associated with the graphics device you want to switch to


## Copying plots

Copying a plot to another device can be useful because some plots require a lot of code and it can be a pain to type all that in again for a different device.

- **dev.copy**: copy a plot from one device to another
- **dev.copy2pdf**: specifically copy a plot to a PDF file

NOTE: Copying a plot is not an exact operation, so the result may not be identical to the origianl
```{r}
library(datasets)
with(faithful, plot(eruptions, waiting)) ## Create a plot on screen device
title(main = "Old Faithful Geyser data") ## Add a main title
dev.copy(png, file = "./support/geyserplot.png") ## Copy my plot to a PNG file
dev.off() ## Don't forget to close the PNG device!
```

## Summary

Plots must be created on a graphics device

The default graphics device is almost always the screen device, which is most useful for exploratory analysis

File devices are useful for creating plots that can be included in other documents or sent to other people

For file devices, there are vector and bitmap formats:

- vector formats are good for line drawings and plots with solid colours using a modest number of points
- bitmap formats are good for plots with a large number of points, natural scenes or web-based plots














